The formal verification of the ctm approach to forcing \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}

### 2022-08-27 Independence_CH/proof_assists_isabelle.tex
Minor typos

To have any value at all, the system should only allow the application
of sound tactics; in particular, neither the author nor (any part) of
the system should be able to fake proofs. Edinburgh LCF~\cite{Gordon1979-qm} was an
influential proof assistant in which the critical code (that
constructs proofs in response to user scripts or other modules) was reduced to a small
\emph{kernel}. Hence, by verfying the correctness of the kernel, one
achieves confidence on the whole system.
%% The rest of the system
%% can interact with the kernel via an interface that exposes functions
%% guaranteeing the validity of the proofs.

@@

Isabelle also inherited from LCF the possibility for the user to define
tactics to encapsulate common patterns used to solve goals. In
fact, this is the origin of the ML family of languages: a
Meta-Language for programming tactics.
In the case of Isabelle, \emph{Standard} ML is the first of the four layers
in which we work in this assistant. Both the kernel and the automation
of proofs are coded in ML, sometimes as a substitute for induction on
formulas, as the next section explains.

@@

For the most part, the development of set theory in Isabelle is
carried out using its ZF object logic
\cite{DBLP:journals/jar/PaulsonG96}, which is the third logical layer
of the formalization and the most versatile one, since 
Isabelle's native automation is available at this level. Apart from
the type and sort
declarations detailed above, it features a finite axiomatization,
% https://isabelle.in.tum.de/dist/library/FOL/FOL/IFOL.html#ZF_Base.mem|const
with a predicate for membership, constants for the empty set and an
infinite set, and functions $\isatt{Pow}::\tyi\fun\tyi$,
$\union::\tyi\fun\tyi$, and $\isatt{PrimReplace} :: \tyi \fun (\tyi
\fun \tyi \fun\tyo)\fun \tyi$ (for Replacement). The Axiom of
Replacement
% https://isabelle.in.tum.de/dist/library/FOL/FOL/IFOL.html#ZF_Base.replacement|axiom
has a free predicate variable $P$: % $P ::\tyi\fun\tyi\fun\tyo$:
\begin{multline*}
  (\forall x \in A .\ \forall y\, z.\ P(x, y) \wedge P(x, z)
  \longrightarrow y=z) \implies \\
  b \in \isatt{PrimReplace}(A, P)
  \longleftrightarrow(\exists x \in A .\ P(x, b)) 
\end{multline*}
The restrictions on sorts described above ensure that it is not
possible that higher-order quantification gets used in $P$. The
statement of $\AC$, also uses a free higher-order variable to denote
an indexed family of nonempty sets. % but we only
% use $\AC$ for demonstration purposes.

>>>>

To have any value at all, the system should only allow the application
of sound tactics; in particular, neither the author nor (any part of)
the system should be able to fake proofs. Edinburgh LCF~\cite{Gordon1979-qm} was an
influential proof assistant in which the critical code (that
constructs proofs in response to user scripts or other modules) was reduced to a small
\emph{kernel}. Hence, by verfying the correctness of the kernel, one
achieves confidence on the whole system.
%% The rest of the system
%% can interact with the kernel via an interface that exposes functions
%% guaranteeing the validity of the proofs.

@@

Isabelle also inherited from LCF the possibility for the user to define
tactics to encapsulate common patterns used to solve goals. In
fact, this is the origin of the ML family of languages: a
Meta-Language for programming tactics.
In the case of Isabelle, \emph{Standard} ML is the first of the four layers
on which we work in this assistant. Both the kernel and the automation
of proofs are coded in ML, sometimes as a substitute for induction on
formulas, as the next section explains.

@@

For the most part, the development of set theory in Isabelle is
carried out using its ZF object logic
\cite{DBLP:journals/jar/PaulsonG96}, which is the third logical layer
of the formalization and the most versatile one, since 
Isabelle's native automation is available at this level. Apart from
the type and sort
declarations detailed above, it features a finite axiomatization,
% https://isabelle.in.tum.de/dist/library/FOL/FOL/IFOL.html#ZF_Base.mem|const
with a predicate for membership, constants for the empty set and an
infinite set, and functions $\isatt{Pow}::\tyi\fun\tyi$,
$\union::\tyi\fun\tyi$, and $\isatt{PrimReplace} :: \tyi \fun (\tyi
\fun \tyi \fun\tyo)\fun \tyi$ (for Replacement). The Axiom of
Replacement
% https://isabelle.in.tum.de/dist/library/FOL/FOL/IFOL.html#ZF_Base.replacement|axiom
has a free predicate variable $P$: % $P ::\tyi\fun\tyi\fun\tyo$:
\begin{multline*}
  (\forall x \in A .\ \forall y\, z.\ P(x, y) \wedge P(x, z)
  \longrightarrow y=z) \implies \\
  b \in \isatt{PrimReplace}(A, P)
  \longleftrightarrow(\exists x \in A .\ P(x, b)) 
\end{multline*}
The restrictions on sorts described above ensure that it is not
possible that higher-order quantification gets used in $P$. The
statement of $\AC$ also uses a free higher-order variable to denote
an indexed family of nonempty sets. % but we only
% use $\AC$ for demonstration purposes.
